Diffusion local times and the time until recombination

### 2015-10-22 notes/diffusion-local-times.Rmd
.

Actually, the cline gets set up
in a time-inhomogeneous way.
Let's do that, by numerical solution,
then run the lineage back through it.
The forwards-time evolution of the cline,
with isotropic density and migration,
solves the PDE (Barton 1979):
$ \begin{align*}
  \partial_t p(x,t) = \frac{\sigma^2}{2} \partial_x^2 + sp(x,t)(1-p(x,t))(2p(x,t)-1) .
\end{align*} $
```{r fwds_pde, cache=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
yinit <- ifelse( thegrid$x.mid>0, 0, 1 )
tt <- seq(0,tau,length.out=100)
fwds.pde <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, D=1/2, dx=thegrid)$dC
    list( tran + s * y * (1-y) * (2*y-1) )
}
fwds.soln <- ode.1D( y=yinit, times=tt, func=fwds.pde, nspec=1 ) # note FIRST COLUMN IS TIME
image(fwds.soln,legend=TRUE,main="mean cline development", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```
Now, let's run the killed lineage backwards in it:
```{r fwds_back_pde, cache=TRUE, depends="fwds_pde"}
yinit <- rep(1.0,thegrid$N)
cline.interp <- function (T) {
    # return the interpolated cline for a particular time
    which.t <- findInterval(T,fwds.soln[,1],rightmost.closed=TRUE)
    lu <- fwds.soln[c(which.t,which.t+1),1]
    alpha <- (T-lu[1])/diff(lu)
    return( (1-alpha)*fwds.soln[which.t,-1] + alpha*fwds.soln[which.t+1,-1]  )
}
rev.pde <- function (t,y,parms,...) {
    p <- cline.interp(tau-t)
    # p <- setup.prop.1D(grid=thegrid, xy=cbind(thegrid$x.mid,cline.interp(tau-t)), interp="linear")
    tran <- tran.1D(C=y, A=p, D=1/2, dx=thegrid)$dC
    list( tran - r * (1-p) * y )
}
rev.soln <- ode.1D( y=yinit, times=tt, func=rev.pde, nspec=1, tcrit=tau ) # note FIRST COLUMN IS TIME
image(rev.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

@@

@@

Here's the same thing for a much longer time:
```{r sim_general_longer, fig.height=fig.dim, cache=TRUE}
nwalks <- 1e3
ntimes <- 1e4
s <- 0.1
tau <- 100
initvals <- c(0,-3)
lt.comparison <- lapply( initvals, function (xinit) { list(
        tanh=weighted_time( 
               sim_diffusion( tanh_drift,tau,nwalks,ntimes,s=s, xinit=xinit ),
               tanh_het, s=s
           ),
        reflected=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                function (x,s) { ifelse(abs(x)<1/(2*sqrt(s)), 1/2, 0) }, s=s
            ),
        reflected_tanh=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                tanh_het, s=s
            ),
        pushed=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                function (x) { x>0 }
            ),
        pushed_tanh=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                tanh_het, s=s
            )
        ) } )
tbreaks <- seq(0,tau,length.out=50)
Tdists <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) tabulate(findInterval(lt,tbreaks),nbins=length(tbreaks)-1) )
    } )
rvals <- exp(seq(log(.001),log(0.5),length.out=400))
precombs <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) sapply( rvals, function (r) { mean((-1)*expm1(-r*lt)) } ) )
    } )
layout(t(1:2))
for (k in seq_along(precombs)) {
    matplot( tbreaks[-1]-diff(tbreaks)/2, Tdists[[k]], type='l', col=1:ncol(Tdists[[k]]), main=paste("x =",initvals[k]), ylim=range(Tdists), xlab="time in hets", lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
```

@@

Here's the cline shape:
```{r show_cline, cache=TRUE}
zeta.pcline <- function (x,r,recombfn=function(x,r){zeta.reflected(x=x/sqrt(tau*sigma^2),r=r*tau)},tau=1,sigma=1) {
    (1/2)*( 1 - sign(x)*zeta.reflected(x=x/sqrt(tau*sigma^2),r=r*tau) )
}
xx <- seq(-15,15,length.out=500)
for (tau in c(5,50)) {
    plot(xx, zeta.pcline(xx,r=1), type='n', 
            xlab="distance from cline center", ylab="prob inherited from the left", 
            main=as.expression(substitute(tau==mytau,list(mytau=tau))) )
    rr <- seq(0,0.2,length.out=11)
    pp <- sapply( rr, function (r) { zeta.pcline(xx,r=r,tau=tau) } )
    matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
    legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
}
```

>>>>

Actually, the cline gets set up
in a time-inhomogeneous way.
Let's do that, by numerical solution,
then run the lineage back through it.
The forwards-time evolution of the cline,
with isotropic density and migration,
solves the PDE (Barton 1979):
$$\begin{aligned*}
  \partial_t p(x,t) = \frac{\sigma^2}{2} \partial_x^2 p(x,t)+ s p(x,t)(1-p(x,t))(2p(x,t)-1) .
\end{aligned*}$$
```{r fwds_pde, cache=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
yinit <- ifelse( thegrid$x.mid>0, 0, 1 )
tt <- seq(0,tau,length.out=100)
fwds.pde <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, D=1/2, dx=thegrid)$dC
    list( tran + s * y * (1-y) * (2*y-1) )
}
fwds.soln <- ode.1D( y=yinit, times=tt, func=fwds.pde, nspec=1 ) # note FIRST COLUMN IS TIME
image(fwds.soln,legend=TRUE,main="mean cline development", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```
Now, let's run the killed lineage backwards in it:
```{r fwds_back_pde, cache=TRUE, depends="fwds_pde"}
yinit <- rep(1.0,thegrid$N)
cline.interp <- function (T) {
    # return the interpolated cline for a particular time
    which.t <- findInterval(T,fwds.soln[,1],rightmost.closed=TRUE)
    lu <- fwds.soln[c(which.t,which.t+1),1]
    alpha <- (T-lu[1])/diff(lu)
    return( (1-alpha)*fwds.soln[which.t,-1] + alpha*fwds.soln[which.t+1,-1]  )
}
rev.pde <- function (t,y,parms,...) {
    p <- cline.interp(tau-t)
    # p <- setup.prop.1D(grid=thegrid, xy=cbind(thegrid$x.mid,cline.interp(tau-t)), interp="linear")
    tran <- tran.1D(C=y, A=p, D=1/2, dx=thegrid)$dC
    list( tran - r * (1-p) * y )
}
rev.soln <- ode.1D( y=yinit, times=tt, func=rev.pde, nspec=1, tcrit=tau ) # note FIRST COLUMN IS TIME
image(rev.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

@@

Just forwards
-------------

An alternate route to this would have been to
solve the coupled PDE for the frequencies of the selected locus 
and the linked locus.

@@

Here's the same thing for a much longer time:
```{r sim_general_longer, fig.height=fig.dim, cache=TRUE}
nwalks <- 1e3
ntimes <- 1e4
s <- 0.1
tau <- 100
initvals <- c(0,-3)
lt.comparison <- lapply( initvals, function (xinit) { list(
        tanh=weighted_time( 
               sim_diffusion( tanh_drift,tau,nwalks,ntimes,s=s, xinit=xinit ),
               tanh_het, s=s
           ),
        reflected=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                function (x,s) { ifelse(abs(x)<1/(2*sqrt(s)), 1/2, 0) }, s=s
            ),
        reflected_tanh=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                tanh_het, s=s
            ),
        pushed=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                function (x) { x>0 }
            ),
        pushed_tanh=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                tanh_het, s=s
            )
        ) } )
tbreaks <- seq(0,tau,length.out=50)
Tdists <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) tabulate(findInterval(lt,tbreaks),nbins=length(tbreaks)-1) )
    } )
rvals <- exp(seq(log(.001),log(0.5),length.out=400))
precombs <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) sapply( rvals, function (r) { mean((-1)*expm1(-r*lt)) } ) )
    } )
layout(t(1:2))
for (k in seq_along(precombs)) {
    matplot( tbreaks[-1]-diff(tbreaks)/2, Tdists[[k]], type='l', col=1:ncol(Tdists[[k]]), main=paste("x =",initvals[k]), ylim=range(Tdists), xlab="time in hets", lty=1 )
    legend("bottomright",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1 )
    legend("bottomright",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
```

@@

Here's the cline shape:
```{r show_cline, cache=TRUE}
zeta.pcline <- function (x,r) { pcline(x,r,recombfn=function(x,r){zeta.reflected(x=x/sqrt(tau*sigma^2),r=r*tau)},tau=1,sigma=1)
xx <- seq(-15,15,length.out=500)
for (tau in c(5,50)) {
    plot(xx, zeta.pcline(xx,r=1), type='n', 
            xlab="distance from cline center", ylab="prob inherited from the left", 
            main=as.expression(substitute(tau==mytau,list(mytau=tau))) )
    rr <- seq(0,0.2,length.out=11)
    pp <- sapply( rr, function (r) { zeta.pcline(xx,r=r,tau=tau) } )
    matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
    legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
}
```
